\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-01-13 doc/publications/plos16/plos16.tex
changes to plos16.tex

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* Denis.Bauer@CSIRO.au

\end{flushleft}
\tableofcontents
\clearpage

\clearpage

@@

Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
large number of areas including variable selection in high dimensions. Consider figure
\ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover the important variables,
plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant variables). There is a
distinct boundary, shown empirically and also based on arguments from combinatorial geometry, to the region where we can
reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behaviour of a number of
regression approaches for variable selection (lars, lasso and forward stepwise) and make the point that above the
phase-transition line variable recovery is still possible by a combinatorial approach.

It is not surprising that that it is more difficult to recover the signal variable in the upper-left area of the figure,
as the problem is both under-determined and sparse. What is surprising is the connection with arguments from
combinatorial geometry. This suggests that we are seeing a universal rule rather than an implementation issue. As
\cursedforest is designed for extremely large numbers of variables it is likely to be operating in difficult regions of
the figure.

We note several things here:
\begin{itemize}
\item the Donoho-Tanner phase transition (in this setting) refers to estimating the $\beta$ in data generated by a
  linear model;
\item in a decision tree (random forest) there is no notion of estimating the $\beta$
\item also a decision tree (random forest) is a heruistic search. It may recover a relationshipin the space
  of the combinatorial search.
\end{itemize}

We note that the examples we are considering lie on the extreme left of the plot, close to the origin. So there are
limits to what can be recovered in the presence of noise. However in the case of data that is both big and wide,
\cursedforest and other VariantSpark methods may provide a useful tool.

>>>>

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* Denis.Bauer@CSIRO.au
\end{flushleft}

\tableofcontents
\clearpage

@@

Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
large number of areas including variable selection in high dimensions. Consider figure
\ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover the important variables,
plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant variables). There is a
distinct boundary, shown empirically and also based on arguments from combinatorial geometry, to the region where we can
reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behaviour of a number of
regression approaches for variable selection (LARS, Lasso and forward stepwise) and make the point that above the
phase-transition line variable recovery is still possible by a combinatorial approach.

It is not surprising that that it is more difficult to recover the signal variable in the upper-left area of the figure,
as the problem is both under-determined and sparse. What is surprising is the connection with arguments from
combinatorial geometry. This suggests that we are seeing a universal rule rather than an implementation issue. As
\cursedforest is designed for extremely large numbers of variables it is likely to be operating in difficult regions of
the figure where the ratio $\delta = n/p$ is small.

We note several things here:
\begin{itemize}
\item the Donoho-Tanner phase transition arises in recovering the $\beta$ in data generated by a
  linear model. However, in a decision tree (random forest) there is no notion of estimating the $\beta$
\item A decision tree (random forest) is a heruistic search. It may recover a relationship in the space of the
  combinatorial search.
\end{itemize}

The existence of the  Donoho-Tanner phase transition is a salutory warning. There are likley to be limits, both
computionaal and logical, to the recovery of signals from noisy data. \cursedforest  is a contribution to addressing the
practivcal limits but the logical limits will still apply. However in the case of data that is both big and wide,
\cursedforest and other VariantSpark methods may provide a useful tool.
